Climatic warming strengthens a positive feedback\\ between alpine shrubs and fire

### 2016-10-18 manuscript/manuscript.Rnw
Updates to manuscript

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b,c]{James S Camac}
\author[d]{Richard J Williams} 
\author[c]{Carl-Henrik Wahren}
\author[e]{Ary A Hoffmann}
\author[b]{Peter A Vesk}

\affil[a]{Department of Biological Sciences, Macquarie University, Sydney, NSW 2109, Australia}
\affil[b]{The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Melbourne, Victoria 3010, Australia}
\affil[c]{Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne, Victoria 3086, Australia}
\affil[d]{CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie, Northern Territory 0821, Australia}
\affil[e]{Bio21 Institute, School of BioSciences, The University of Melbourne, Victoria 3010, Australia}

@@

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, it is unknown whether increases in woody species and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen fire-killed alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that warming doubled tall shrub seedling growth rates and could increase survival. Warming had limited or no effect on shrub recruitment, post-fire gap-infilling by grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore biotic interactions and the type, frequency and severity of disturbances that are also likely to change \citep{Guisan:2005gz,Tucker:2012dr,Giannini:2013gi}. Because disturbance strongly influences recruitment opportunities, and thus, composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}, it has the capacity to exacerbate or diminish how climate affects vegetation. Seedlings are the life stage most vulnerable to climate and disturbance, and the life stage which determines the long-term persistence of a species and its capacity to establish in new areas. Consequently, it is imperative to understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv}.

Field manipulative experiments have been widely used to investigate the effects of climate change in alpine and tundra plant communities. These experiments have provided invaluable information on how climate directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}. However, because the general perception is that alpine and tundra vegetation is predominately clonal \citep{Billings:1968bn}, despite contradictory evidence \citep[e.g.][]{Venn:2009bw,Briceno:2015iv}, coupled with the fact that disturbances vary both spatially and temporally, most studies have focused on mature plant responses in relatively undisturbed vegetation \citep{Briceno:2015iv}. As such, few field climate experiments have been conducted on seedlings or vegetation in post-disturbance conditions \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}. Consequently, little is known about how climate change influences seedling demographic rates \citep{Briceno:2015iv} and thus, vegetation recovery in post-disturbance conditions.

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ua, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover \citep[e.g. undisturbed alpine tussock grassland;][]{Williams:1987ua}. Evidence also indicates that shrubs can have strong supressive effects on other life forms \citep{MyersSmith:2011fp}, that climatic warming is likely to increase woody species growth rates \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp}, and that shrubs are the most flammable component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. The combination of this evidence indicates that more frequent or severe fire may increase recruitment opportunities for woody species, which when coupled with higher growth and survival rates caused by rising temperatures, will exacerbate shrub expansion \citep{Racine:2004tn,Lantz:2013fh} and ultimately increase the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. In effect, warming could strengthen an existing feedback loop that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback loop. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid lines = known mechansims; Dashed lines = mechansims that may exacerbate or diminish this feedback loop but which we have a paucity of information on. (b) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dashed line) that could break this feedback. Blue = positive relationships, Red = negative relationships.}
\label{fig1}
\end{figure}

While paleoecological studies have indicated that such a feedback may exist \citep{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

We examine these unknowns and their impacts on the feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with tundra ecosystems, the Australian Alps have experienced rapid changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent wildfires, the frequency and severity of which are expected to increase \citep{Bradstock:2014jp,Williams:2014ud}.

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To determine drivers of shrub establishment and how warmer post-fire conditions affect shrub vital rates, we combined field observations with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub). This combination of data sources allowed us to examine: 1) landscape scale drivers of shrub seedling establishment; 2) how warmer post-fire conditions influence rates of grass recovery and shrub seedling recruitment, growth and mortality; 3) how proximity to grass affect shrub seedling demographic rates; and 4) whether such grass effects change under warmer post-fire conditions.

\section*{Methods}
We investigated shrub dynamics in open heathland because it is a common and highly flammable plant community in the Australian Alps above 1600 m a.s.l \citep{Williams:2006tl}. It is also an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland consisting of \textit{Grevillea australis} shrubs interspersed among a \textit{Poa hiemata} sward. Aerial photography has revealed that the plant community has acted as an invasion front of shrub expansion into grassland and has itself experienced shrub thickening by closed heathland dominants such as \textit{Prostanthera cuneata} \citep{McDougall:2003gb}.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5 m$^2$) \textit{Grevillea australis} shrubs in open heathland to create patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance in open heathland burnt by wildfire. After creating the bare ground patches we collected seedlings of dominant alpine shrub species from a nearby (\textless 2 km) site of similar altitude burnt by a late 2006 wildfire. We collected shrub seedlings of two dominant open heathland species \textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland \textit{Phebalium squamulosum} (Rutaceae; a tall shrub). We focused on these four species because they are all fire-killed \citep[the dominant shrub post-fire strategy in the Australian mainland Alps;][]{Walsh:2004vk}; common in the Australian Alps; and under climatic warming, are likely candidates to invading non-shrubby plant communities such as alpine grasslands and herbfields.

A total of 640 seedlings, 256 \textit{Grevillea} (half used in \textit{Poa} inter-tussock experiment\textemdash see below) and 128 for each of \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} were used. Four seedlings per species were randomly selected and transplanted into a 4\texttimes 4 square grid in the center of each burnt patch, with 14 cm between individuals and the edge of the patch dominated by resprouting tussock grass \textit{Poa hiemata}. To examine interactions between tussock grass and shrub seedlings we also randomly transplanted four additional \textit{Grevillea australis} into various sized inter-tussock gaps between burnt \textit{Poa hiemata} immediately surrounding the bare ground (Fig. \ref{figs8}). The experimental site was fenced to prevent grazing by deer and horses. We detected no obvious signs of rabbit or invertebrate herbivory within our plots.

To simulate near-term warmer conditions indicated by the \citet*{IPCC:2013vv}, we randomly assigned Open Top Chambers (OTCs) to half (16) the plots, with the remainder treated as unwarmed controls. OTCs were placed over plots, ensuring all seedlings (including inter-tussock shrub seedlings) occurred within the 1.1 m$^2$ open top to minimise edge effects. OTCs were placed out at the start of the growing season (October) where they remained until snowfall (early June). This procedure was repeated for five growing seasons from May 2010 to May 2015.

Microclimatic conditions were measured hourly using Onset Micro Stations (Onset Computer Corporation, Bourne, MA, USA) at four control and four OTC plots. Across \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} growing seasons (\Sexpr{n_days(chamber_dates, TRUE)} growing season days), OTCs simulated warmer conditions at the lower end of \citet*{IPCC:2013vv} projections (Fig. \ref{figs5}-\ref{figs7}). OTCs passively increased average ambient and soil temperatures by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C. Minimum and maximum temperatures were also raised in both ambient air (min: \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean_min','grand_mean')}\textdegree C; max: \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean_max','grand_mean')}\textdegree C) and soil (min: \Sexpr{micro_clim(overall_microclimate_summary, 'CTemp_3cmBG','mean_min','grand_mean')}\textdegree C; max: \Sexpr{micro_clim(overall_microclimate_summary, 'CTemp_3cmBG','mean_max','grand_mean')}\textdegree C). Chambers only marginally decreased soil moisture by \Sexpr{abs(micro_clim(overall_microclimate_summary, 'CMoisture3to10cmBG','mean','grand_mean'))}\% and relative humidity by \Sexpr{abs(micro_clim(overall_microclimate_summary, 'CRH','mean','grand_mean'))}\%.

Seedling survival, maximum height and stem diameter (nearest mm measured with Vernier calipers) were initially recorded in May 2010 and then subsequently re-measured at the end of each growing season (May-June). At the same time, we recorded the distance to the nearest tussock or grass seedling in each of four cardinal directions for shrub seedlings growing in inter-tussock gaps. Thus, changes in inter-tussock gap size could be due to either vegetative growth or seedling establishment. We did not measure individual characteristics (e.g. height and basal diameter) of surrounding tussock grass because we could not distinguish individuals, and because height varied throughout the season. Numbers of natural \textit{Grevillea australis} recruits establishing within the plots were also recorded for the first two seasons.

@@

We used 40 long-term open heathland sites established after the 2003 fires \citep{Camac:2013eb} consisting of 30 burnt sites and 10 sites thought to be unburnt for over 70 years. The disparity in sampling effort between burnt and unburnt was due to very few open heathland remaining unburnt after both the 2003 and 2007 fires. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index \citep[TWI; a measure of plant available water,][]{Moore:1993fk} were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by twig diameters, collected immediately after the 2003 fires \citep{Camac:2013eb}.

@@

In total we built 11 hierarchical models to examine factors influencing post-fire recovery of tussock grass and shrub seedling establishment, growth and mortality. For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package \texttt{rstan} \Sexpr{packageVersion("rstan")}. Data and source code for reproducing analysis and figures are available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. Additional information about experimental design and analysis can be found in Supporting Information.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[a,b]{James S Camac}
\author[c]{Richard J Williams} 
\author[b]{Carl-Henrik Wahren}
\author[d]{Ary A Hoffmann}
\author[a]{Peter A Vesk}

\affil[a]{The Centre of Excellence for Environmental Decisions, School of BioSciences, The University of Melbourne, Melbourne, Victoria 3010, Australia}
\affil[b]{Research Centre for Applied Alpine Ecology, La Trobe University, Melbourne, Victoria 3086, Australia}
\affil[c]{CSIRO Land and Water, Tropical Ecosystems Research Centre, PMB 44 Winnellie, Northern Territory 0821, Australia}
\affil[d]{Bio21 Institute, School of BioSciences, The University of Melbourne, Victoria 3010, Australia}

@@

\begin{abstract}
Climate change is expected to increase fire activity and woody plant encroachment in arctic and alpine landscapes. However, the extent to which these increases interact to affect changes in structure, function and composition of alpine ecosystems is unknown. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of four dominant, evergreen fire-killed alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}\textendash fold) and that experimental warming doubled tall shrub seedling growth rates and could potentially increase survival. Warming had no effect on shrub recruitment, post-fire infilling of gaps by a dominant alpine grass, or competitive effects of grass on shrub seedling growth and survival. These findings indicate that warming, coupled with more frequent or severe fires, will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing shrub survival. Thus, warming is likely to enhance a positive feedback between woody species abundance and fire activity in alpine and tundra environments.
\end{abstract}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\section*{Introduction}
Accurately forecasting the effects of climatic warming on vegetation dynamics requires an understanding of the mechanisms by which climate and vegetation interact. Most forecasting models include the direct effects of climatic conditions on species distributions, but largely ignore disturbances, the type, frequency and severity of which are also likely to change \citep{Guisan:2005gz,Tucker:2012dr}. Recurrent disturbance is integral to all ecosystems, strongly influencing fundamental demographic processes such as recruitment and mortality, and thus the composition and structure of plant communities and biomes worldwide \citep{Bond:2005eg, Turner:2010wo}. Disturbance regimes are likely to change as a consequence of climate change \citep{Turner:2010wo, Westerling:2011jp,Bradstock:2014jp}. Moreover, the effects of disturbance will interact with climate change, and thus may amplify or diminish how changes to temperature and moisture affect vegetation \citep{Post:2008wz,Camac:2015hja,Enright:2015em}

Feedbacks between climate change, vegetation and fire will be particularly important determinants of the trajectory of ecosystem change in coming decades \citep{Bowman:2009kp}. For example, in addition to the direct effects of warming on fire weather, changing climate may also alter fire regimes by altering vegetation productivity, structure and plant trait composition \citep{Keeley:2012wo,Matthews:2012gia,Bowman:2014dr}. Feedbacks may be positive or negative, and have been documented in tropical savanna \citep{Hoffmann:2003br}, boreal forests \citep{Krawchuk:2011wr}, Mediterranean ecosystems \citep{Keeley:2012wo} and grasslands \citep{Flannigan:2009kl}. Change in both vegetation and fire regimes in alpine and arctic environments are also likely to be subject to feedbacks \citep{Goetz:2007gj,Wookey:2009wo}.

Alpine and arctic vegetation are particularly vulnerable to the effects of changing climate \citep{Engler:2011vq,Dullinger:2012cg, Elmendorf:2012kz}. Field manipulative experiments have been widely used to investigate the potential effects of climate change in alpine and tundra plant communities. These experiments have provided invaluable information on how climate directly influences plant phenology \citep{Hoffmann:2010kf, Dorji:2013tt}, reproduction \citep{Klady:2011dw}, morphology \citep{Hudson:2009cn}, growth \citep{Hollister:2005vf}, floristic composition \citep{Elmendorf:2012kz} and biotic interactions \citep{Klanderud:2005uw}.

However, most studies have focused on plant responses in mature, undisturbed vegetation \citep{Briceno:2015iv}. Few climate-change field experiments have included disturbance as a factor \citep[but see:][]{Munier:2010cg,Graae:2010ir,Camac:2015hja}, or examined the influence of climate change on seedlings in post-disturbance environments. Although alpine and tundra vegetation has a strong capacity to regenerate vegetatively \citep{Billings:1968bn} there is mounting evidence that seedling regeneration is important in these ecosystems \citep{Venn:2009bw, Briceno:2015iv}, particularly for wood species \citep{Camac:2013eb,Williams:2014ud}. Although seedlings are highly vulnerable to climate and disturbance, they are the life stage which determines the long-term persistence of a species and its capacity to establish in new areas. However, little is known about how climate change affects seedling demography \citep{Briceno:2015iv}, particularly in post-disturbance conditions in alpine and tundra ecosystems. It is therefore imperative we understand how seedlings respond to both changing climate and disturbance regimes \citep{Walck:2011dm, Briceno:2015iv} if we wish to accurately predict future trajectories of vegetation change.

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species \citep{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue \citep{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas \citep{Williams:1987ws, Batllori:2009fe, Frost:2013tm} and lowest in areas with minimal bare ground cover \citep[e.g. undisturbed alpine tussock grassland;][]{Williams:1987ws}. Current evidence indicates that climatic warming is likely to increase woody species growth rates \citep{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp} and that shrubs are the most flammable vegetation component of these ecosystems \citep{Williams:2006tl, Higuera:2009ga, Fraser:NHT5Kx_W}. Thus, increases in the cover of woody species in alpine and arctic environments is likely to increase the flammability of these ecosystems.

The combination of this evidence indicates a potential positive feedback between warmer temperatures, woody species dynamics, and fire activity in alpine and tundra environments (Fig. \ref{fig1}a). Warmer temperatures are likely to lead to more frequent or severe fire. This is likely to increase woody species recruitment opportunities (i.e. more bare ground) and exacerbate both shrub thickening within shrub communities and shrub expansion into shrub-free communities such as grassland and herbfields \citep{Racine:2004tn,Lantz:2013fh}. If this effect is coupled with increased shurb seedling growth and mortality, highly flammable fuels will accumulate faster and have higher landscape connectivity, ultimately leading to further increases in the likelihood of fire \citep{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. Thus, warming could strengthen an existing climate-disturbance feedback, that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also has serious social, biodiversity and carbon sequestration consequences \citep{Mack:2011ew}.

\begin{figure}[!h]
\includegraphics{figures/fig1.pdf}
\caption{Conceptual diagram illustrating positive warming-shrub-fire feedback. (a) Hypothesised positive feedback loop between fire, climatic warming and landscape flammability. Solid lines = known mechansims; Dashed lines = mechansims that may exacerbate or diminish this feedback but which we have a paucity of information on. (b) Feedback loop based on new evidence obtained from experimental and field surveys conducted in this study. Also includes an example of a potential, but unlikely, effect (red dashed line) that could break this feedback. Blue = positive relationships, Red = negative relationships.}
\label{fig1}
\end{figure}

While paleoecological studies have indicated that such a feedback may exist in the arctic \citep{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on what may strengthen, mitigate or break this feedback. For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer, more exposed, post-fire conditions. Furthermore, we do not know how tussock grasslands (a dominant structural vegetation type in arctic and alpine ecosystems) will respond to warmer post-fire conditions, whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

Here, we examine these unknowns and their impacts on this proposed feedback between climatic warming, shrubs and fire in Australian alpine vegetation. As with other tundra ecosystems, Australian alpine landscapes consist of a matrix of grass-herb dominanted vegetation types, and shrub-dominanted types \citep{Williams:2014ud}. They have also experienced significant changes in climate. Since 1979, mean growing season temperatures have risen by approximately 0.4\textdegree C and annual precipitation has fallen by 6\% \citep{Wahren:2013ji}, with a consequent decline in snow pack depth \citep{SanchezBayo:2013ci}. These climatic changes have been correlated with a 10 to 20\% increase in shrub cover and a 25\% decline in graminoids cover \citep{Wahren:2013ji}. Much of the Australian Alps has also been burnt by recent (2003 and 2006) wildfires, the frequency and/or severity of which are expected to increase as a consequence of climate change \citep{Bradstock:2014jp,Williams:2014ud}.

We took advantage of recent fires in alpine open heathland, a plant community that occupies ca. 25\% of the Australian alpine landscape \citep{Williams:2014ud}. Under global warming, this plant community is likely to encroach upon grasslands, wetlands and herbfields \citep[including the nationally endangered snow patch herbfields;][]{Williams:2015gj}, and is itself susceptible to shrub thickening \citep{McDougall:2003gb}. To determine drivers of shrub establishment and how warmer post-fire conditions affect shrub vital rates, we combined field observations of post-fire shrub seedling density with a warming experiment that used seedlings of four dominant Australian evergreen fire-killed shrubs: \textit{Grevillea australis} (Proteaceae; a tall shrub), \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub),  \textit{Phebalium squamulosum} (Rutaceae; a tall shrub) and \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub).

This combination of data sources allowed us to quantify some of the outstanding unknowns in the hypothesised climate-fire-shrub feedback loop (Fig. \ref{fig1}a). Specifically, we examined: 
\begin{enumerate}\itemsep 0pt
  \item Landscape scale drivers of shrub seedling establishment.
  \item How warmer post-fire conditions influence rates of grass recovery and shrub seedling recruitment, growth and mortality.
  \item How proximity to grass affects shrub seedling demographic rates; and whether such grass effects change under warmer post-fire conditions.
\end{enumerate}

\section*{Methods}

A summary of study sites, experimental design and approach to data analysis is given below. Full details are presented in the Supporting Information.

\subsection*{Study Sites}

The Australian Alps are an ideal place to examine potential fire-climate-vegetation feedbacks in high mountain environments because Australian alpine ecosystems have all the elements needed to demonstrate such potential feedbacks. Here, the Alps are subject to recurrent landscape fire, and were burnt extensively in 2003 \citep{Williams:2006tl,Williams:2014ud}. The vegetation consists of a range of life forms (graminoids, other herbs, shrubs), and variation in life-form dominance between and within vegetation types is a strong determinant of landscape flammability \citep{Williams:2006tl,Fraser:NHT5Kx_W}. There is also evidence that both fire regimes and the relative abundance of lifeforms in the Australian Alps have changed as a consequence of recent climate change \citep{Bradstock:2014jp,Wahren:2013ji,Camac:2015hja}.

We investigated shrub-grass dynamics in open heathland vegetation on the Bogong High Plains, Australia. Open heathland is a common Australian alpine vegetation type, occupying approximately 25\% of the treeless landscapes above 1600m in the Australian Alps. It is dominated by the shrub \textit{Grevillea australis} (shrub cover generally 20\textendash 60\%) with inter-shrub regions occupied by snow grasses (\textit{Poa hiemata} and \textit{Poa costiniana}) and other herbaceous species. It is an ecotone between closed heathland (\textgreater 70\% shrub cover) and tussock grassland (shrub cover < 20\%). \textit{Grevillea australis} is an obligate seeding shrub, and establishment of seedlings is dependent on disturbance creating bare ground in the grass\textendash herb sward \citep{Williams:1987ws}. Relative to tussock grassland, open heathland is a highly flammable plant community \citep{Williams:2006tl,Fraser:NHT5Kx_W}. A consequence of this differential flammability lead to the 2003 wildfires on the Bogong High Plains burnting approximately 60\% of open heathland and only 13\% of tussock grasslands. Following these fires, seedling regeneration of \textit{Grevillea australis} and other shrubs was prolific following the 2003 fires \citep{Williams:2014ud}.

@@

In March 2010, at 1750 m a.s.l, we burnt 32 randomly selected mature (60 cm tall and 1.5 m$^2$) \textit{Grevillea australis} shrubs in an open heathland site that was not burnt by wildfire in 2003 or 2006. This created patches of bare ground approximately 0.7 m$^2$ surrounded by burnt tussock grass, simulating disturbance of individual shrubs in open heathland burnt by wildfire. We collected seedlings of dominant alpine shrub species from a nearby (\textless 2 km) site of similar altitude burnt by 2006 wildfires. We collected seedlings of two dominant open heathland shrub species \textit{Grevillea australis} (Proteaceae; a tall shrub) and \textit{Asterolasia trymalioides} (Rutaceae; a prostrate shrub), a dominant closed heathland species \textit{Prostanthera cuneata} (Lamiaceae; a tall shrub) that typically grows on warmer aspects and a species common to both open and closed heathland \textit{Phebalium squamulosum} (Rutaceae; a tall shrub). We focused on these four species because they are common in the Australian Alps; are fire killed and thereby re-establish via seed \citep[the dominant shrub post-fire strategy in the Australian mainland Alps;][]{Walsh:2004vk}; and under climatic warming, have the potential to increase in cover and height within heathlands, and invade non-shrubby plant communities such as alpine grasslands and herbfields.

A total of 640 seedlings, 256 \textit{Grevillea} (half used in \textit{Poa} inter-tussock experiment\textemdash see below) and 128 for each of \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} were used. Four seedlings per species were randomly selected and transplanted into a 4\texttimes 4 square grid in the center of each burnt patch, with 14 cm between individuals and the edge of the patch, which was dominated by resprouting tussock grass \textit{Poa hiemata}. To examine interactions between tussock grass and shrub seedlings we also randomly transplanted four additional \textit{Grevillea australis} seedlings into various sized bare gaps between burnt \textit{Poa hiemata} tussocks that were immediately surrounding the 4\texttimes 4 bare ground square grid (Fig. \ref{figs1}). The experimental site was fenced to prevent grazing by deer and horses. We detected no obvious signs of herbivory by invertebrates or hares within our plots.

To simulate near-term warmer conditions indicated by the \citet*{IPCC:2013vv}, we randomly assigned Open Top Chambers (OTCs) to half (16) the plots, with the remainder treated as unwarmed controls. The chambers were constructed following the International Tundra Experiment (ITEX) protocols \citep{Molau:1996vc}. OTCs were placed over plots, ensuring all seedlings (including inter-tussock shrub seedlings) occurred within the 1.1 m$^2$ open top to minimise edge effects. OTCs were placed out at the start of the growing season (October) where they remained until first snowfall (early June). This procedure was repeated for six growing seasons from May 2010 to May 2016.

Microclimatic conditions were measured hourly using Onset Micro Stations (Onset Computer Corporation, Bourne, MA, USA) at four control and four OTC plots. Across \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} growing seasons (\Sexpr{n_days(chamber_dates, TRUE)} growing season days), OTCs simulated warmer conditions at the lower end of \citet*{IPCC:2013vv} projections (Fig. \ref{figs6}-\ref{figs8}). OTCs passively increased average ambient and soil temperatures by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C. Minimum and maximum temperatures were also raised in both ambient air (min: \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean_min','grand_mean')}\textdegree C; max: \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean_max','grand_mean')}\textdegree C) and soil (min: \Sexpr{micro_clim(overall_microclimate_summary, 'CTemp_3cmBG','mean_min','grand_mean')}\textdegree C; max: \Sexpr{micro_clim(overall_microclimate_summary, 'CTemp_3cmBG','mean_max','grand_mean')}\textdegree C). Chambers only marginally decreased soil moisture by \Sexpr{abs(micro_clim(overall_microclimate_summary, 'CMoisture3to10cmBG','mean','grand_mean'))}\% and relative humidity by \Sexpr{abs(micro_clim(overall_microclimate_summary, 'CRH','mean','grand_mean'))}\%.

Seedling survival, maximum height and stem diameter (nearest mm measured with Vernier calipers) were initially recorded in May 2010 and then subsequently re-measured at the end of each growing season (May-June). At the same time, we recorded the distance to the nearest tussock or grass seedling in each of four cardinal directions for \textit{Grevillea} seedlings growing in inter-tussock gaps. Thus, temporal changes in inter-tussock gap size could be due to either vegetative growth of resprouting tussocks, or the establishment of grass seedlings. We did not measure individual characteristics (e.g. height and basal diameter) of surrounding tussock grass because we could not distinguish individuals, and because height varied substantially throughout the season. Numbers of natural \textit{Grevillea australis} recruits establishing within the plots were also recorded and identified for the first two years of the experiment.

@@

We used 40 open heathland monitoring sites established after the 2003 fires \citep{Williams:2006tl,Camac:2013eb}. These sites consisted of 30 burnt sites and 10 sites known to be unburnt for over 70 years. The disparity in sampling effort between burnt and unburnt was due to very few open heathland sites of sufficient size (0.25 ha) remaining unburnt after both the 2003 and 2006 fires. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 quadrats, each 1 m$^2$, that were evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persisted post-fire. We were unable to estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index \citep[TWI; a measure of plant available water,][]{Moore:1993fk} were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated by twig diameters \citep{Whight:1999vw}, collected immediately after the 2003 fires \citep{Williams:2006tl}.

@@

We built multiple hierarchical models to examine how increased temperature and other factors influenced shrub seedling recruitment, growth and mortality, as well as tussock-grass gap dynamics. For each model we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} \citep{Rcran:2016xx} using package \texttt{rstan} \Sexpr{packageVersion("rstan")} \citep{stan:2015xx}. Detailed information about experimental design and analysis is provided in the Supporting Information. Data and source code are available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. In order to aid in the reproducibility of this work, our code was written using a \texttt{remake} framework \citep{fitzjohn2015}, such that others can readily reproduce our entire workflow from data processing, through to producing a pdf of this manuscript by calling \texttt{remake()} in R.
